---
title: "Sleep Study"
author: "Johanna Wahn"
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_chunk$set(fig.align="center")
library(png)
library(grid)
img <- readPNG("IMG_3053.png")
```

## Why we are doing this study?
```{r Pomi Image, echo=F}
grid.raster(img)
```

*This is a study case used as a capstone for the Google Data Analytics certificate*

**Problem**: My favorite activity in the world to do with my cat is sleeping. Could I optimize it by controlling the different variables that might affect it?


**Solution**: Let's dive into a sleeping study to discover which variables will improve or ruin a good night sleep!

## Statistical Information

```{r Load Packages, echo=T, results='hide'}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(patchwork)
library(ggpubr)
library(GGally)
library(caret)
```

```{r Read Data}
data <- read.csv("sleepdata.csv")
head(data)
```

```{r }
summary(data)
```

```{r null values}
#Number of null values per column
colSums(is.na(data))
```

```{r Unique IDs}
length(unique(data$Person.ID)) == nrow(data)
```

```{r average age and BMI per gender}
gen_avg <- data %>% 
  group_by(Gender) %>%
  summarise(mean_age = mean(Age), n = n())

gen_avg
```
The average age of Women is higher. This could skew the data as age could be an important factor in sleep quality. But we do have a 50/50 gender representation.

## Exploratory Data Analysis (EDA)

``` {r Exploratory graphs}
data$Sleep.Disorder = factor(data$Sleep.Disorder, levels = c('None','Insomnia','Sleep Apnea'))

df <- data %>% 
  group_by(Sleep.Disorder) %>% # Variable to be transformed
  count() %>% 
  ungroup() %>% 
  mutate(perc = `n` / sum(`n`)) %>% 
  arrange(perc) %>%
  mutate(labels = scales::percent(perc))

ggplot(df, aes(x = "", y = perc, fill = Sleep.Disorder)) +
  geom_col() +
  geom_label(aes(label = labels), color = "black",
            position = position_stack(vjust = 0.5),
            show.legend = FALSE) +
  scale_fill_brewer(palette = "Reds") +  
  coord_polar("y", start = 0) +  
  theme_void() +
  ggtitle("Percentage of Sleep Disorders") +
  theme(plot.title = element_text(hjust=0.5))


```

We do not have an equal representation of every sleep class, but we almost have a 60/40 representation of healthy and disease respectively.

```{r Quality of Sleep}
my_comparisons <- list( c("None", "Insomnia"), 
                        c("Insomnia", "Sleep Apnea"), 
                        c("None", "Sleep Apnea") )

ggplot(data, aes(x=Sleep.Disorder, y=Quality.of.Sleep, fill=Sleep.Disorder)) +
  geom_boxplot()+
  geom_jitter(shape=16, position=position_jitter(0.2)) +
  scale_fill_brewer(palette="Reds") +
  stat_summary(fun = "mean", geom = "point", shape = 8,
               size = 2, color = "white")+ 
  stat_compare_means(comparisons = my_comparisons)+ # Add pairwise comparisons p-value
  stat_compare_means(label.y = 12) + # Add global p-value
  ggtitle("Quality fo Sleep score distribution \n per Sleep Disorder") +
  theme(plot.title = element_text(hjust=0.5))     
```

The average quality of aleep score for a person without a disorder is ~7.5. Sleep apnea does not seem to be correlated with a lower quality of sleep, we can see no significant difference between the Sleep Apnea and No Disorder classes. Insomnia is the only class where we see a significant loss in quality of sleep. The quality of sleep metric seems to be very subjective as it does not definitely indicate a sleep disorder or not.

```{r Gender}

gend_group <- data %>%
  group_by(Gender,Sleep.Disorder,Age)

gender_bar <- ggplot(data, aes(x=Gender, y=Age, fill=Sleep.Disorder)) +
  scale_fill_brewer(palette="Reds") +
  geom_bar(stat="summary", fun.y = "mean", position="dodge") +
  ggtitle("Sleep Disorders per Gender")+
  theme(plot.title = element_text(hjust=0.5))     

gender_bar 
```

As stated before, the female group has a higher average age and seems more affected by sleep apnea that the male group, while the male group seems the most affected by insomnia. 

```{r BMI}
data$BMI.Category <- gsub("Normal Weight", "Normal", data$BMI.Category)
data$BMI.Category <- factor(data$BMI.Category, labels=c('Normal','Overweight','Obese'))

bmi_bar <- ggplot(data, aes(x=BMI.Category, fill=Sleep.Disorder)) +
  scale_fill_brewer(palette="Reds") +
  geom_bar() +
  ggtitle("BMI effect on Sleep") +
  theme(plot.title = element_text(hjust=0.5))
bmi_bar
```

We cleaned up this column by merging "Normal" and "Normal Weight" samples. The overweight category seems severely under represented. 

A conclusion we can pull from this figure is that BMI is correlated with sleep quality. A higher BMI tends to indicate insomnia and/or sleep apnea, while in the lowest BMI category has a majority of normal sleep. Obese class has almost 50/50 split on both sleep disorders. We cannot make an conclusions on the Overweight category due to the small sample size but it does seem to indicate a correlation with sleep disorders.

```{r Occupation and Stress, fig.width=12, fig.height=6}
data_stress <- data %>%
  group_by(Occupation) %>%
  summarize(avg_stress = mean(Stress.Level),
            avg_age = mean(Age),
            count = n())
  
occup_point <- ggplot(data, aes(x=Occupation, y=Stress.Level)) +
  geom_point(aes(fill=Sleep.Disorder,size=Age), color='black', shape=21, stroke=0.4) +
  scale_fill_brewer(palette="Reds") +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=0.5),
        plot.margin = margin(t = 5, r = 10, b = 5, l = 10),
        axis.title.x = element_text(margin = margin(t = 20)),
        plot.title = element_text(hjust=0.5)) +
  ggtitle("Occupation, Age and Stress \n effect on Sleep") 

occup_bar <- ggplot(data, aes(x=Occupation, fill=Sleep.Disorder)) +
  geom_bar() +
  scale_fill_brewer(palette="Reds")+
  ggtitle("Occupation \n effect on Sleep") +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=0.7),
        plot.margin = margin(t = 5, r = 10, b = 5, l = 10),
        axis.title.x = element_text(margin = margin(t = 20)),
        plot.title = element_text(hjust=0.5)) 

print(data_stress)

occup_bar + occup_point + plot_layout(widths = c(1, 1))
```
The occupations with the highest stress scores seem to be the most correlated with sleep disorders, as illustrated by Sales representative, Salespersons and Scientists, especially causing sleep apnea. The occupations with a lower average stress score but a higher age group average, like nurses and teachers, also show an increase in the number of sleep disorders, even when the stress score is on the low end (see Nurses with score of 3).

```{r Blood ranges}
BP_ranges = c('Normal', 'Elevated', 'Hypertension_1', 'Hypertension_2', 'Hypertensive_Crisis')
BP_systolic_limits = list(c(0,120),c(120,130),c(130,140),c(140,180),c(180,200))
BP_diastolic_limits = list(c(0,80),c(0,80),c(80,90),c(90,120),c(120,140))

# Categorize into BP ranges depending on the systolic and diastolic values 
data <- data %>%
  separate(Blood.Pressure, into=c('Systolic','Diastolic'), sep="/") %>%
  mutate(Systolic = as.numeric(Systolic),
         Diastolic = as.numeric(Diastolic)) %>%
  mutate(BP_range = case_when(
    between(Systolic, BP_systolic_limits[[1]][1], BP_systolic_limits[[1]][2]) & 
      between(Diastolic, BP_diastolic_limits[[1]][1], BP_diastolic_limits[[1]][2]) ~ BP_ranges[1],
    
    between(Systolic, BP_systolic_limits[[2]][1], BP_systolic_limits[[2]][2]) & 
      between(Diastolic, BP_diastolic_limits[[2]][1], BP_diastolic_limits[[2]][2]) ~ BP_ranges[2],
    
    between(Systolic, BP_systolic_limits[[3]][1], BP_systolic_limits[[3]][2]) & 
      between(Diastolic, BP_diastolic_limits[[3]][1], BP_diastolic_limits[[3]][2]) ~ BP_ranges[3],
    
    between(Systolic, BP_systolic_limits[[4]][1], BP_systolic_limits[[4]][2]) & 
      between(Diastolic, BP_diastolic_limits[[4]][1], BP_diastolic_limits[[4]][2]) ~ BP_ranges[4],
    
    between(Systolic, BP_systolic_limits[[5]][1], BP_systolic_limits[[5]][2]) & 
      between(Diastolic, BP_diastolic_limits[[5]][1], BP_diastolic_limits[[5]][2]) ~ BP_ranges[5],
    
    TRUE ~ "Unknown" # Default case
  ))


# New values to replace "Unknown"
replacement_values <- c(rep("Elevated", 12), rep("Hypertension_1", 2), "Elevated")

data <- data %>%
  mutate(BP_range = replace(BP_range, BP_range == "Unknown", replacement_values)) %>%
  mutate(Heart.Rate = as.factor(Heart.Rate))

data$Heart.Rate <- as.numeric(as.character(data$Heart.Rate))
# Group by BP_range and Heart.Rate, then count occurrences
data_grouped <- data %>%
  group_by(BP_range, Heart.Rate, Sleep.Disorder, Age) %>%
  summarise(Count = n(), .groups = "drop")  # Count occurrences of each Heart Rate

# Convert BP_range into a factor with the specified order
data_grouped$BP_range <- factor(data_grouped$BP_range, levels = BP_ranges)

blood_bar <- ggplot(data_grouped, aes(x = BP_range, y = Count, fill = Heart.Rate)) +
  geom_col() +  
  scale_fill_viridis_c(option = "magma", direction = -1) +  
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  ggtitle("Heart rate of \n of blood pressure ranges") +
  theme(plot.title = element_text(hjust=0.5))

blood_bar
```

We separated the "Blood.Pressure" column into "Systolic" (#/) and "Diastolic" (/#) categories and then depending on their combination create a categorical blood pressure range column: "BP_range" from the [American Heart Association](https://www.heart.org/en/health-topics/high-blood-pressure/understanding-blood-pressure-readings) . We manually clean up the "Unknown" ranges by making assumptions (putting more importance on the systolic number) and assigning them a category.

As expected, we can see an increase in heart rate with each blood pressure range. 

```{r BP and Heart Rate}
data_grouped$Sleep.Disorder = factor(data_grouped$Sleep.Disorder, levels = 
                                       c('None','Insomnia','Sleep Apnea'))

blood_sleep <- ggplot(data_grouped, aes(x=BP_range, y=Heart.Rate)) + 
  geom_jitter(aes(fill=Sleep.Disorder,size=Age), color='black', 
              shape=21, stroke=0.4, width=0.3) + 
  scale_fill_brewer(palette="Reds") +
  ggtitle("Effect of BP range, Heart Rate \n and Age on Sleep") +
  theme(plot.title = element_text(hjust=0.5))

blood_sleep
```
We see an evident correlation between hypertension and sleep disorders. Regardless of age, a lower heart rate and a normal BP range indicates a lower propensity to sleep disorders, while the hypertension I and II have a higher proportion of insomnia and sleep apnea, with hypertension II being almost exclusively being sleep apnea. An elevated BP does not seem correlated with sleep disorders.

```{r Sport and Stress, fig.width=12, fig.height=6}
data <- data %>% 
  mutate(Stress.Level = as.factor(Stress.Level))

act_bar <- ggplot(data, aes(x=Physical.Activity.Level,y=Daily.Steps)) +
  geom_jitter(aes(size=Age, fill=Stress.Level), color='black', shape=21, stroke=0.4) +
  scale_fill_brewer(palette="Blues") +
  geom_smooth(method=lm,  linetype="dashed",
              color="darkblue") +
  ggtitle("Effect of Physical Activity, \n Daily Steps and Age on Stress") +
  theme(plot.title = element_text(hjust=0.5))

act_stress_bar <- ggplot(data, aes(x=Physical.Activity.Level,y=Daily.Steps)) +
  geom_jitter(aes(fill=Sleep.Disorder, size=Age), color='black', shape=21, stroke=0.4) +
  scale_fill_brewer(palette="Reds") +
  geom_smooth(method=lm,  linetype="dashed",
              color="darkred")+
  ggtitle("Effect of Physical Activity, \n Daily Steps and Age on Sleep") +
  theme(plot.title = element_text(hjust=0.5))

act_bar + act_stress_bar + plot_layout(widths = c(1, 1))
```

We see a really strong correlation between physical activity and daily steps with the stress level, the lower the physical activity and daily steps, the higher the stress level the person feels regardless of their age. We see a couple of outliers where the stress levels are high at the top right (highest number of steps with highest activity level). As seen by the following table:

```{r Stress, Activity and Steps}
data %>%
  group_by(Occupation, Stress.Level) %>%
  summarise(avg_steps = mean(Daily.Steps),
            avg_activity = mean(Physical.Activity.Level))
```

The higher number of steps could be a result of the job and not a result of leisure sport, especially with nurses who have the highest number of steps over all the other occupations (8058 steps).

Insomnia seems to be correlated with a lower level of physical, while sleep apnea does not follow this pattern. The latter one seems to mainly affect older people regardless of the physical activity.

```{r Sleep duration and quality}

sleep_q <- ggplot(data, aes(x=Sleep.Duration,y=Quality.of.Sleep)) +
  geom_jitter(aes(fill=Sleep.Disorder), color='black', 
              shape=21, stroke=0.4, size=3, width=0.5) +  
  scale_fill_brewer(palette="Reds") +
  ggtitle("Effect of Sleep Duration and \n Quality of Sleep on Sleep Disorders") +
  theme(plot.title = element_text(hjust=0.5))

sleep_q
```

Quality of sleep seems to be strongly correlated with sleep duration. As expected insomnia is correlated with a lower sleep duration and quality of sleep. Sleep apnea does not seem to be correlated with neither sleep duration nor sleep quality, we find sleep anea in both high and low quality of sleep score ranges as well as short and long sleep duration. As quality of sleep is a subjective measure if the sleep apnea is mild it might not affect the person itself (maybe only those around them lol).

```{r Correlation}
library(corrplot)

# Compute correlation matrix (only for numeric variables)
numeric_data <- data[sapply(data, is.numeric)]
numeric_data <- numeric_data[2:ncol(numeric_data)]
cor_matrix <- cor(numeric_data, use="complete.obs")

# Visualize using corrplot
corrplot(cor_matrix, method="color", type="upper", 
         col= colorRampPalette(c("blue", "white", "red"))(200),
         tl.col="black", tl.srt=45)  # Adjust label angle
```

## Conclusions 

We have a variety of variables affecting sleep disorders. Some of them are inevitable such as age or gender. Older folks and women seem to have a higher correlation with sleep apnea and insomnia. However some of the variables we have a say over, such as BMI, physical activity, stress levels and occupation (maybe this last one is not so easy a change). All of the aforementioned variables are also strongly correlated with each other. Higher physical activity also means lower stress and blood pressure ,in addition we could hypothesize a lower BMI. Improving all these variables would inversely correlate with sleep disorders. 

**The goal of this study case was to use data analytics to derive insights, however I want to also apply some machine learning to obtain an order of feature importance and to see if this data is enough to predict a sleep disorder and to push myself to perform some ML in R instead of my usual python.**

## Pre-processing data

- Unique IDs will be deleted
- Gender will become a binary variable
- Ordinal Categorical values will be transformed to integers (Sleep.Disorder)
- Nominal Categorical values (Occupation, BMI.Category, BP_range) will be one shot encoded or label encoded
- Blood.Pressure will be divided into systolic and diastolic values
- The rest of the values will be standardized as needed

```{r ML preprocessing}
df <- data %>%
  select(!c(Person.ID)) %>%
  mutate(
    Gender = if_else(data$Gender == "Female", 0, 1), #Female=0, Male=1,
    Occupation = as.integer(factor(data$Occupation)),
    BMI.Category = as.integer(factor(data$BMI.Category)),
    BP_range = as.integer(factor(BP_range)),
    Stress.Level = as.integer(Stress.Level)
  )

```

We now check the distributions of the numeric variables that will maybe need to be standardized: Sleep Duration, Quality of Sleep, Physical Activity, 

```{r Check distributions 1, fig.width=12, fig.height=6}

age <- ggplot(data, aes(x=Age)) + geom_histogram() 
sleepdur <- ggplot(data, aes(x=Sleep.Duration)) + geom_histogram() 
sleepqual <- ggplot(data, aes(x=Quality.of.Sleep)) + geom_histogram() 

age+sleepdur+sleepqual+ plot_layout(widths = c(1, 1,1))

```

```{r Check distributions 2, fig.width=12, fig.height=6}

phy <- ggplot(data, aes(x=Physical.Activity.Level)) + geom_histogram() 
stress <- ggplot(data, aes(x=Stress.Level)) + geom_histogram(stat="count") 
sys <- ggplot(data, aes(x=Systolic)) + geom_histogram() 

phy+stress+sys+plot_layout(widths = c(1, 1,1))
```

```{r Check distributions 3, fig.width=12, fig.height=6}

dia <- ggplot(data, aes(x=Diastolic)) + geom_histogram() 
hr <- ggplot(data, aes(x=Heart.Rate)) + geom_histogram() 
ds <- ggplot(data, aes(x=Daily.Steps)) + geom_histogram() 

dia+hr+ds+ plot_layout(widths = c(1, 1,1))
```

We have really uneven distributions for the data, but due to the small number of samples I do not want to manipulate the data too much and skew the results. We will use a random forrest model that does not require standardisation of the input data.

```{r Train/Test split}
set.seed(123)  # For reproducibility
split <- createDataPartition(df$Sleep.Disorder, p = 0.8, list = FALSE)

train_data <- df[split, ]
test_data  <- df[-split, ]
```

```{r Random Forrest Model}
# Convert target variable to factor (for classification)
train_data$Sleep.Disorder <- as.factor(train_data$Sleep.Disorder)
test_data$Sleep.Disorder  <- as.factor(test_data$Sleep.Disorder)

# Set up training control with cross-validation
train_control <- trainControl(method = "cv", number = 5)  

# Train Random Forest model
model <- train(Sleep.Disorder ~ ., data = train_data,
               method = "rf",  
               trControl = train_control,
               tuneLength = 5)  

# View model details
print(model)

# Make predictions
predictions <- predict(model, newdata = test_data)

# Evaluate accuracy
conf_matrix <- confusionMatrix(predictions, test_data$Sleep.Disorder)
print(conf_matrix)
```


## Conclusions 

The model can predict a sleep disorder with a 93% accuracy. We do note an imbalance in sensitivity between the different groups but as seeing as there is a very imbalanced number of samples for each group this was to be expected. 

```{r Pomi Image 2, echo=F}
img <- readPNG("pom_reading.png")
grid.raster(img)
```

Dr. Pomelo revising my script.